Estimating SARS-CoV-2 transmission in England from randomly sampled prevalence surveys

### 2023-10-27 ATACCC.tex
Edit notation

@@

@@

For an individual $i$, their trajectory is determined by two points and one slope parameter.
The points are the time at which the individual has a 50\% chance of testing positive, $t_{i,0}$, and the time at which their viral load
peaks, $t_{i,\text{peak}}$ after $t_{i,0}$; the slope parameter, $\beta_i$ determines the rate their Ct increases (viral load decreases) following the peak.
The Ct value for an individual at time $t_{i,0}$ is the known limit of detection, $C_\text{lod}$,
Their Ct value at the peak is $C_{i,\text{peak}}$ and inferred.
The slope parameter, $\beta_i$, determines how quickly an individual's Ct value increases following the peak viral load.
We denote the vector of the parameters defined here as $\vec\theta_i = \begin{bmatrix} t_{i,0} & t_{i,\text{peak}} & C_{i,\text{peak}} &  \beta_i \end{bmatrix}^T$.

This parameterisation leads to the following Ct value trajectory for individual $i$:
$$
\hat{C}_i(t, \vec\theta_i) = \begin{cases}
  C_{\text{lod}} + (C_{i,\text{peak}} - C_{\text{lod}}) \frac{t - t_{i,0}}{t_{i,\text{peak}}}
    &t \leq t_{i,\text{peak}} + t_{i,0} \\
  C_{i,\text{peak}} + \beta_i (t - t_{i,\text{peak}} - t_{i,0})
    &t > t_{i,\text{peak}} + t_{i,0}.
\end{cases}
$$

@@

Observed Ct values are assumed to be drawn from a Normal distribution with mean $\hat{C}_i(t)$ and standard deviation $\sigma_\text{obs}$.
censored at the limit of detection.
Additionally, we introduce a probability of a false negative $\eta$, to account for negative test results seen near the peak viral load.
Denoting an observed Ct value for individual $i$ at time $t$ as $y_{i,t}$ with
$y_{i,t} = \infty$ for a negative test, this gives the following likelihood.
$$
\pi(y_{i,t} \mid \vec\theta_i, \sigma_\text{obs}, \eta) = \begin{cases}
  (1 - \eta) f_N(y_{i,t} \mid \hat{C}_i(t, \vec\theta_i), \sigma_\text{obs}^2) &y_{i,t} < \infty \\
  \eta + (1 - \eta) (1 - F_N(C_\text{lod} \mid \hat{C}_i(t, \vec\theta_i), \sigma_\text{obs}^2)) & y_{i,t} = \infty
\end{cases}
$$
where $\sigma_\text{obs}$ is a model parameter determining the variability in
test results, and $f_N(y \mid \mu, \sigma^2)$ and $F_N(y \mid \mu, \sigma^2)$ are the pdf
and cdf (respectively) of a Normal distribution with mean $\mu$ and variance $\sigma^2$.

@@

% Define the nodes
\node[state] (mu-tpeak) {$\mu_{\log t_\text{peak}}$};
\node[state, right=of mu-tpeak] (sigma-logtpeak) {$\sigma_{\log t_\text{peak}}$};
\node[state, below right=of mu-tpeak] (ti-peak) {$t_{i,\text{peak}}$};
\node[state, right=of sigma-logtpeak] (mu-Cpeak) {$\mu_{C_\text{peak}}$};
\node[state, right=of mu-Cpeak] (sigma-Cpeak) {$\sigma_{C_\text{peak}}$};
\node[state, below right=of mu-Cpeak] (Ci-peak) {$C_{i,\text{peak}}$};
\node[state, right=of sigma-Cpeak] (mu-log-Bi) {$\mu_{\log \beta_i}$};
\node[state, right=of mu-log-Bi] (sigma-log-Bi) {$\sigma_{\log \beta_i}$};
\node[state, below right=of mu-log-Bi] (Bi) {$\beta_i$};
\node[state, left=of ti-peak] (ti0) {$t_{i,0}$};
\node[draw, inner sep=0.3cm, fit=(ti0) (Bi), fill=blue!10,opacity=0.2] {$\theta_i$};
\node[state, below left=of Ci-peak] (Ci-t) {$\hat{C}_i(t)$};
\node[state, right=of Ci-t] (sigma-obs) {$\sigma_\text{obs}$};
\node[state, right=of sigma-obs] (eta-i) {$\eta_i$};
\node[rectangle, draw, left=of Ci-t,minimum size=1.5cm] (Clod) {$C_\text{lod}$};
\node[state, double, below=of sigma-obs] (Yi-t) {$y_{i,t}$};

% Draw the edges
\draw[->] (mu-tpeak) -- (ti-peak);
\draw[->] (sigma-logtpeak) -- (ti-peak);
\draw[->] (mu-Cpeak) -- (Ci-peak);
\draw[->] (sigma-Cpeak) -- (Ci-peak);
\draw[->] (mu-log-Bi) -- (Bi);
\draw[->] (sigma-log-Bi) -- (Bi);
\draw[->,dotted] (ti0) -- (Ci-t);
\draw[->,dotted] (ti-peak) -- (Ci-t);
\draw[->,dotted] (Ci-peak) -- (Ci-t);
\draw[->,dotted] (Bi) -- (Ci-t);
%\draw[->] (theta-i) -- (Ci-t);
\draw[->] (Ci-t) -- (Yi-t);
\draw[->] (eta-i) -- (Yi-t);
\draw[->,dotted] (Clod) -- (Ci-t);
\draw[->] (Clod) -- (Yi-t);
\draw[->] (sigma-obs) -- (Yi-t);

\end{tikzpicture}
\caption{Hierarchical structure}
\end{figure}

@@

\begin{table}[ht]
\centering
\begin{tabular}{llrrr}
  \hline
    Parameter & Distribution & 0.05 & 0.5 & 0.95 \\ 
  \hline
    $\eta$ & $\text{Beta}(0.95, 9)$ & $0$ & $0.07$ & $0.28$ \\ 
    $\sigma_\text{obs}$ & $\text{Half-Cauchy}(0, 0.67)$ & $0.05$ & $0.67$ & $8.51$ \\ 
    $t_{0,i}$ & $N(0, 5^2)$ & $-8.22$ & $0$ & $8.22$ \\ 
    $\mu_{\log t_\text{peak}}$ & $N(\log(3), \log(2.5)^2)$ & $\log (0.66)$ & $\log (3)$ & $\log (13.54)$ \\ 
    $\sigma_{\log t_\text{peak}}$ & $\text{Half-Cauchy}(0, 0.67)$ & $0.05$ & $0.67$ & $8.51$ \\ 
    $\mu_{C_\text{peak}}$ & $N(17, 5^2)$ & $8.78$ & $17$ & $25.22$ \\ 
    $\sigma_{C_\text{peak}}$ & $\text{Half-Cauchy}(0, 4)$ & $0.31$ & $4$ & $50.82$ \\ 
    $\mu_{\log\beta}$ & $N(\log(3.5), \log(1.4)^2)$ & $\log (2.01)$ & $\log (3.5)$ & $\log (6.09)$ \\ 
    $\sigma_{\log\beta}$ & $\text{Half-Cauchy}(0, 0.67)$ & $0.05$ & $0.67$ & $8.51$ \\
   \hline
\end{tabular}
\caption{Priors used, columns 3 to 5 gives quantiles of the distributions specified. See main text for description of the parameters.}
\label{tab:paper:priors}
\end{table}

@@

>>>>

Overall summary of chapter.
Need to include: how the different sections and analyses interact.

\Cref{ATACCC:sec:background} explains how biomarkers are typically modelled, both for SARS-CoV-2 and historically in the biostatistical literature.
I explain an analysis, \textcite{hakkiOnset}, applying this to the ATACCC data in \cref{ATACCC:sec:hakki}.
However, this analysis did not converge fully.
I modify their observation model in section \cref{ATACCC:sec:observation-modification} to improve the convergence.
This modification is based on an earlier analysis I did with a slightly different model and a preliminary version of the dataset, which is described in \cref{ATACCC:sec:original-analysis}.
The model and parameter estimates of the biomarker need to be converted to a duration of PCR positivity to be used in backcalculation, which I do in \cref{ATACCC:sec:duration}.

@@

This section introduces the piecewise linear model I initially used to model viral loads.
It consists of two steps:...

@@

For an individual $i$, their trajectory is determined by two points and one slope parameter.
The points are the time at which the individual has a 50\% chance of testing positive, $t_{i,0}$, and the time at which their viral load peaks relative to their first swab, $t_{i,\text{peak}}$ after $t_{i,0}$; the slope parameter, $\beta_i$ determines the rate their viral load decreases following the peak.
The viral load value for an individual at time $t_{i,0}$ is the known limit of detection, $v_\text{lod}$,
Their viral load value value at the peak is $v_{i,\text{peak}}$ and inferred.
The slope parameter, $\beta_i$, determines how quickly an individual's viral load decreases following the peak viral load.
We denote the vector of the parameters defined here as $\vec\theta_i = \begin{bmatrix} t_{i,0} & t_{i,\text{peak}} & v_{i,\text{peak}} &  \beta_i \end{bmatrix}^T$.

This parameterisation leads to the following value for individual $i$'s log viral load:
$$
\hat{v}_i(t, \vec\theta_i) = \begin{cases}
  v_{\text{lod}} + (v_{i,\text{peak}} - v_{\text{lod}}) \frac{t - t_{i,0}}{t_{i,\text{peak}}}
    &t \leq t_{i,\text{peak}} + t_{i,0} \\
  v_{i,\text{peak}} - \beta_i (t - t_{i,\text{peak}} - t_{i,0})
    &t > t_{i,\text{peak}} + t_{i,0}.
\end{cases}
$$

@@

Observed log viral load values are assumed to be drawn from a Normal distribution with mean $\hat{v}_i(t)$ and standard deviation $\sigma_\text{obs}$, censored at the limit of detection.
Additionally, we introduce a probability of a false negative $\eta$, to account for negative test results seen near the peak viral load.
Denoting an observed log viral load value for individual $i$ at time $t$ as $y_{i,t}$ with
$y_{i,t} = \infty$ for a negative test, this gives the following likelihood.
$$
\pi(y_{i,t} \mid \vec\theta_i, \sigma_\text{obs}, \eta) = \begin{cases}
  (1 - \eta) f_N(y_{i,t} \mid \hat{v}_i(t, \vec\theta_i), \sigma_\text{obs}^2) &y_{i,t} < \infty \\
  \eta + (1 - \eta) F_N(v_\text{lod} \mid \hat{v}_i(t, \vec\theta_i), \sigma_\text{obs}^2) & y_{i,t} = \infty
\end{cases}
$$
where $\sigma_\text{obs}$ is a model parameter determining the variability in
test results, and $f_N(y \mid \mu, \sigma^2)$ and $F_N(y \mid \mu, \sigma^2)$ are the pdf
and cdf (respectively) of a Normal distribution with mean $\mu$ and variance $\sigma^2$.

@@

% Define the nodes
\node[state] (mu-tpeak) {$\mu_{\log t_\text{peak}}$};
\node[state, right=of mu-tpeak] (sigma-logtpeak) {$\sigma_{\log t_\text{peak}}$};
\node[state, below right=of mu-tpeak] (ti-peak) {$t_{i,\text{peak}}$};
\node[state, right=of sigma-logtpeak] (mu-vpeak) {$\mu_{v_\text{peak}}$};
\node[state, right=of mu-vpeak] (sigma-vpeak) {$\sigma_{v_\text{peak}}$};
\node[state, below right=of mu-vpeak] (vi-peak) {$v_{i,\text{peak}}$};
\node[state, right=of sigma-vpeak] (mu-log-Bi) {$\mu_{\log \beta_i}$};
\node[state, right=of mu-log-Bi] (sigma-log-Bi) {$\sigma_{\log \beta_i}$};
\node[state, below right=of mu-log-Bi] (Bi) {$\beta_i$};
\node[state, left=of ti-peak] (ti0) {$t_{i,0}$};
\node[draw, inner sep=0.3cm, fit=(ti0) (Bi), fill=blue!10,opacity=0.2] {$\theta_i$};
\node[state, below left=of vi-peak] (vi-t) {$\hat{v}_i(t)$};
\node[state, right=of vi-t] (sigma-obs) {$\sigma_\text{obs}$};
\node[state, right=of sigma-obs] (eta-i) {$\eta_i$};
\node[rectangle, draw, left=of vi-t,minimum size=1.5cm] (vlod) {$v_\text{lod}$};
\node[state, double, below=of sigma-obs] (Yi-t) {$y_{i,t}$};

% Draw the edges
\draw[->] (mu-tpeak) -- (ti-peak);
\draw[->] (sigma-logtpeak) -- (ti-peak);
\draw[->] (mu-vpeak) -- (vi-peak);
\draw[->] (sigma-vpeak) -- (vi-peak);
\draw[->] (mu-log-Bi) -- (Bi);
\draw[->] (sigma-log-Bi) -- (Bi);
\draw[->,dotted] (ti0) -- (vi-t);
\draw[->,dotted] (ti-peak) -- (vi-t);
\draw[->,dotted] (vi-peak) -- (vi-t);
\draw[->,dotted] (Bi) -- (vi-t);
%\draw[->] (theta-i) -- (vi-t);
\draw[->] (vi-t) -- (Yi-t);
\draw[->] (eta-i) -- (Yi-t);
\draw[->,dotted] (vlod) -- (vi-t);
\draw[->] (vlod) -- (Yi-t);
\draw[->] (sigma-obs) -- (Yi-t);

\end{tikzpicture}
\caption[Hierarchical structure]{Hierarchical structure}
\end{figure}

@@

\begin{table}[ht]
\centering
\begin{tabular}{llrrr}
  \hline
    Parameter & Distribution & 0.05 & 0.5 & 0.95 \\ 
  \hline
    $\eta$ & $\text{Beta}(0.95, 9)$ & $0$ & $0.07$ & $0.28$ \\ 
    $\sigma_\text{obs}$ & $\text{Half-Cauchy}(0, 0.67)$ & $0.05$ & $0.67$ & $8.51$ \\ 
    $t_{0,i}$ & $N(0, 5^2)$ & $-8.22$ & $0$ & $8.22$ \\ 
    $\mu_{\log t_\text{peak}}$ & $N(\log(3), \log(2.5)^2)$ & $\log (0.66)$ & $\log (3)$ & $\log (13.54)$ \\ 
    $\sigma_{\log t_\text{peak}}$ & $\text{Half-Cauchy}(0, 0.67)$ & $0.05$ & $0.67$ & $8.51$ \\ 
    $\mu_{v_\text{peak}}$ & $N(17, 5^2)$ & $8.78$ & $17$ & $25.22$ \\ 
    $\sigma_{v_\text{peak}}$ & $\text{Half-Cauchy}(0, 4)$ & $0.31$ & $4$ & $50.82$ \\ 
    $\mu_{\log\beta}$ & $N(\log(3.5), \log(1.4)^2)$ & $\log (2.01)$ & $\log (3.5)$ & $\log (6.09)$ \\ 
    $\sigma_{\log\beta}$ & $\text{Half-Cauchy}(0, 0.67)$ & $0.05$ & $0.67$ & $8.51$ \\
   \hline
\end{tabular}
\caption{Priors used, columns 3 to 5 gives quantiles of the distributions specified. See main text for description of the parameters.}
\label{tab:paper:priors}
\end{table}

@@

\subsection{Viral load model}

\subsection{Observation model}

\subsection{Within-individual correlation}
